---
title: "DA 401 Early Results"
author: "Calvin Burzynski"
date: "2025-10-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
library(tidyr)
library(stringr)
library(lubridate)
library(readr)
library(ggplot2)
library(fixest)
library(janitor)
library(broom)

#load in ipums (acs) dataset
ipums <- read.csv("../../../usa_00003.csv.gz")

#load minimum wage DOL data
min_wage_data <- read.csv("..//..//..//Minimum Wage from DOL 2018-2024.csv")

#load monthly unemployment rate data
unemployment_rate_data <- read.csv("..//..//..//State unemployment rates over the last 10 years, seasonally adjusted_BLS.csv")

gdp_income <- read.csv("..//..//..//State Annual Summary Statistics GDP Income.csv", skip = 3)
```


```{r}
#reusable visual theme for graphs 
theme_slide <- theme_minimal(base_size = 13) + 
  theme(
    plot.title.position = "plot",
    plot.title = element_text(face = "bold"), 
    plot.subtitle = element_text(margin = margin(b = 10)), 
    plot.caption = element_text(size = 9, color = "grey40"), 
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.title.y = element_text(margin = margin(r = 8)),
    axis.title.x = element_text(margin = margin(t = 8)),
    legend.position = "bottom"
  )
```


```{r}
#filter for poverty status
ipums %>%
  distinct(POVERTY) %>%
  arrange(POVERTY) #investigating values of poverty 

ipums_poverty <- ipums %>%
  filter(EMPSTAT == 1, UHRSWORK >= 35, WKSWORK2 >= 6, AGE >= 18, AGE <= 64) %>%
  mutate(
    in_poverty = ifelse(OFFPOV == 1, 1, 0), 
    valid = ifelse(OFFPOV %in% c(0, 1), 1, 0)
  ) %>%
  filter(valid == 1) %>%
  group_by(STATEFIP, YEAR) %>%
  summarise(
    poverty_rate = weighted.mean(in_poverty, w = PERWT, na.rm = TRUE),
    pop_weight = sum(PERWT, na.rm = TRUE), 
    .groups = 'drop'
  )
#pop_weight represents the estimated total number of individuals represented by the ACS sample in that state-year 


```

```{r}
min_wage_long <- min_wage_data %>%
  rename_with(~str_remove(., "^X"), starts_with("X")) %>% #cleans columns with 'X' prefix
  pivot_longer(
    cols = all_of(as.character(2018:2024)),
    names_to = "YEAR",
    values_to = "min_wage"
  ) %>%
  mutate(
    YEAR = as.numeric(YEAR)
  ) %>%
  filter(between(YEAR, 2018, 2023))
```

```{r}
#pivot unemplyoment rate data to long format

#replaces "." with "-" for month and year columns
unemployment_rate_rename <- unemployment_rate_data %>%
  rename_with(~ str_replace_all(.x, "\\.", "-"), .cols = -State)

#helps match structure of month-year columns 
month_cols <- grep("^[A-Za-z]{3}-\\d{2,4}$", names(unemployment_rate_rename), value = TRUE)

unemployment_numeric <- unemployment_rate_rename %>%
  mutate(across(all_of(month_cols), ~ parse_number(as.character(.))))

unemployment_long <- unemployment_numeric %>%
  pivot_longer(
    cols = all_of(month_cols), 
    names_to = "date_string", 
    values_to = "unemployment_rate"
  ) %>%
  mutate(
    date = my(date_string), 
    YEAR = year(date)
  ) %>%
  filter(!is.na(date), between(YEAR, 2018, 2023)) %>%
  group_by(State, YEAR) %>%
  summarise(unemployment_rate = mean(unemployment_rate, na.rm = TRUE), .groups = 'drop')
```

```{r}
gdp_income_long <- gdp_income %>%
  rename_with(~str_remove(., "^X"), starts_with("X")) %>%
  filter(LineCode %in% c(1, 8)) %>%
  select(GeoName, GeoFips, LineCode, all_of(as.character(2018:2023))) %>%
  pivot_longer(
    cols = all_of(as.character(2018:2023)),
    names_to = "YEAR",
    values_to = "value"
  ) %>%
  mutate(
    YEAR = as.integer(YEAR), 
    value = parse_number(as.character(value)),
    STATEFIP = as.integer(parse_number(as.character(GeoFips)) / 1000L), 
    var = case_when(
      LineCode == 1 ~ "real_gdp", 
      LineCode == 8 ~ "real_per_capita_income"
    )
  ) %>%
  filter(STATEFIP >0) %>%
  select(STATEFIP, YEAR, var, value, GeoName) %>%
  pivot_wider(names_from = var, values_from = value) %>%
  mutate(real_gdp = as.numeric(real_gdp), 
         real_per_capita_income = as.numeric(real_per_capita_income)
         )
```

```{r}
unemployment_with_fips <- unemployment_long %>%
  left_join(
    min_wage_long %>%
      distinct(State, StateFIPS) %>%
      transmute(State, STATEFIP = as.integer(StateFIPS)), 
    by = "State"
  ) %>%
  transmute(STATEFIP, YEAR, unemployment_rate)
```

```{r}
#merging datasets
panel_data <- ipums_poverty %>%
  left_join(min_wage_long %>% transmute(STATEFIP = as.integer(StateFIPS), YEAR, min_wage), 
            by = c("STATEFIP", "YEAR")) %>%
  left_join(unemployment_with_fips, by = c("STATEFIP", "YEAR")) %>%
  left_join(gdp_income_long, by = c("STATEFIP", "YEAR"))


#filter out district of columbia
panel_clean <- panel_data %>%
  filter(STATEFIP != 11)

#summary of data
summary(panel_clean)
names(panel_clean)
```

```{r}
#check data distributions

#gdp 
ggplot(panel_clean, aes(x = real_gdp)) + geom_histogram(bins = 30, fill = "darkgreen", color = "black") + labs(title = "Distribution of GDP")

#income
ggplot(panel_clean, aes(x = real_per_capita_income)) + geom_histogram(bins = 30, fill = "darkblue", color = "white") + labs(title = "Distribution of Real Per Capita Income")

#minimum wage 
ggplot(panel_clean, aes(x = min_wage)) + geom_histogram(bins = 30, fill = "darkred", color = "white") + labs(title = "Distribution of Minimum Wage Across States")


#due to skewness of variables and common practice in research, some log transformations will be applied

#also creating dummy variable and log of one that looks okay distribution wise, but can be used down the line for robustness check
panel_clean2 <- panel_clean %>%
  mutate(log_gdp = log(pmax(real_gdp, 1)), 
         log_income = log(pmax(real_per_capita_income, 1)), 
         log_minw = log(pmax(min_wage, 1)), 
         treated_minw = as.integer(min_wage > 7.25))
  
```

```{r}
plot_data <- panel_clean2 %>%
  group_by(STATEFIP) %>%
  mutate(Group = ifelse(any(min_wage > 7.25), "Treated", "Control")) %>%
  ungroup()


ggplot(plot_data, aes(x = as.numeric(as.character(YEAR)), y = poverty_rate, color = Group)) + 
  stat_summary(fun = "mean", geom = "line", size = 1.2) + 
  stat_summary(fun = "mean", geom = "point", size = 2.5) + 
  labs(title = "Parallel Trends Check: Treated vs. Control States",
       subtitle = "Average poverty rates before and after minimum wage increases",
       x = "Year", y = "Mean Povery Rate", color = "Group") + 
  theme_minimal()
```

The parallel trends visualization above compares the average poverty rates between treated states (those with minimum wages above \$7.25) and control states (those maintaining the federal minimum) from 2018 to 2023. Prior to policy changes, both groups appear to exhibit stable but distinct trajectories. This results in the treated states showing consistently lower poverty levels, while control states experience slightly higher rates. Although the two groups are not perfectly parallel in the pre-treatment period, their trends move in broadly similar directions, especially before 2020, suggesting that parallel trends assumption is reasonably plausible. The post-2020 divergence, particularly the sharper increase among control states, is consistent with the hypothesis that minimum wage increases may have mitigated poverty growth in treated states following the economic disruptions of Covid-19. Moving from 2022 to 2023, we can see the treatment and control group start to follow similar trends.

```{r}
ggplot(panel_clean2, aes(x = poverty_rate)) + geom_histogram(bins = 30, fill = "blue", color = "black") + labs(title = "Distribution of State-Year Poverty Rates")


```

```{r}
#poverty vs minimum wage
ggplot(panel_clean2, aes(x = log_minw, y = poverty_rate)) + 
  geom_point(alpha = 0.6, color = "blue") + 
  geom_smooth(method = "lm", se = FALSE, color = "red", size = 1) + 
  labs(title = "Poverty vs. Minimum Wage", 
       x = "Log(Minimum Wage)", 
       y = "Poverty Rate") + theme_slide
```

The scatterplot shows a negative relationship between poverty rates and real per capita income across states and years. As income levels rise, poverty rates consistently decline, which aligns with economic theory that higher income improves household well-being and reduces the share of individuals living below the poverty threshold. The fitted regression line highlights a moderately strong downward slope, suggesting that even small proportional increases in per capita income are associated with measurable reductions in state-level poverty. This relationship provides early support for including real per capita income as a key control variable in assessing the effects of minimum wage policy on poverty.


```{r}
#poverty vs. income
ggplot(panel_clean2, aes(x = log_income, y = poverty_rate)) +
  geom_point(alpha = 0.6, color = "black") + geom_smooth(method = "lm", se = FALSE, color = "orange", size = 1) + 
  labs(title = "Poverty vs Real Per Capita Income", x = "Log(Income)", y = "Poverty Rate") + theme_slide
```

The relationship visualized between poverty rates and the log of minimum wage was also negative. though somewhat weaker and more dispersed than that observed for income. States with higher minimum wages generally exhibit lower poverty rates, but the wider scatter reflects that wage policy alone does not fully explain poverty variation across states. Still, the downward-sloping trend line indicates that increases in the minimum wage are correlated with lower poverty, supporting the hypothesis that higher wage floors can improve living standards for low-income workers. The visualization provides early evidence of the expected direction of the policy's impact. which will be tested more formally in the fixed-effects regressions. 

```{r}
panel_clean2 %>%
  group_by(treated_minw) %>%
  summarise(
    mean_poverty = mean(poverty_rate, na.rm = T),
    mean_minw = mean(min_wage, na.rm = T),
    mean_income = mean(real_per_capita_income, na.rm = T),
    mean_unemp = mean(unemployment_rate, na.rm = T)
  )
```


```{r}
str(panel_clean2)
names(panel_clean2)

#function "feols" is having issues with finding STATEFIP even though it is in the dataset
vars_needed <- c("STATEFIP", "YEAR", "poverty_rate", "log_minw", "log_income", "log_gdp",
                 "unemployment_rate", "pop_weight")
setdiff(vars_needed, names(panel_clean2)) #returned character(0) so no actual issues with names

#convert to factors
panel_clean2$YEAR <- as.factor(as.numeric(panel_clean2$YEAR))
panel_clean2$STATEFIP <- as.factor(as.numeric(panel_clean2$YEAR))

fe_model <- feols(
  poverty_rate ~ log_minw + log_income + log_gdp + unemployment_rate | STATEFIP + YEAR,
  data = panel_clean2,
  cluster = ~ STATEFIP,
  weights = ~ pop_weight,
)
summary(fe_model)

#polished results of fixed effects regression
etable(fe_model, digits = 3, title = "Fixed Effects Model: Poverty vs. Minimum Wage")
```

The fixed-effects regression results indicate that, after accounting for state and year differences as well as macroeconomic controls, increases in the minimum wage are associated with statistically significant reductions in poverty. Specifically, the coefficient on log (minimum wage) is -0.012 and significant at the 0.1% level, implying that a 1% rise in the minimum wage corresponds to roughly a 0.012 percentage decline in the poverty rate, holding other factors constant. Both log (income per capita) (-0.048) and log (GDP) (0.001) are significant, with the income coefficient confirming that higher household income substantially reduces poverty, while GDP's small positive coefficient suggests that aggregate growth alone does not necessarily translate into immediate poverty alleviation at the household level. The unemployment rate also shows a small but significant positive association with poverty, consistent with expectations that higher unemployment elevates economic hardship. With a within R-squared of 0.57, the model explains over half of the variation in state-level poverty across time, underscoring that while minimum wage policy has a measurable effect, labor market conditions and income growth remain the dominant influences on poverty reduction between 2018 and 2023.  

```{r}
coef_data <- tidy(fe_model, conf.int = T)

coef_data %>%
  filter(term == "log_minw") %>%
  ggplot(aes(x = term, y = estimate)) + 
  geom_point(color = "darkred", size = 3) + 
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2, color = "gray40") + 
  labs(title = "Effect of Log (Minimum Wage) on Poverty Rate", 
       x = "", y = "Estimated Coefficient") + theme_minimal()
``` 

The coefficient plot displays the estimated relationship between the log of minimum wage and the state-level poverty rate from the fixed effects regression. The red dot represents the point estimate, while the gray vertical line shows the 95% confidence interval. The estimate of approximately -0.012 indicates that, while holding constant real per capita income, real GDP, and unemployment, a 1% increase in the minimum wage is associated with roughly a 0.012 percentage point decrease in the poverty rate. Although the direction is negative

```{r}
coef_keep <- c("log_minw", "log_income", "log_gdp", "unemployment_rate")

coef_df <- tidy(fe_model, conf.int = T) %>%
  filter(term %in% coef_keep) %>%
  mutate(term = factor(term, 
                       levels = rev(coef_keep), 
                       labels = rev(c("Log (Min Wage)", "Log (Income Per Capita)", "Log (Real GDP)", "Unemployment Rate"))))

p_coef <- ggplot(coef_df, aes(x = term, y = estimate)) + 
  geom_hline(yintercept = 0, linetype = "dashed") + 
  geom_pointrange(aes(ymin = conf.los, ymax = conf.high), size = 0.8) + 
  coord_flip() + labs(
    title = "Fixed Effects Model Coefficients", 
    subtitle = "State & Year FE; clustered by state; pop-weighted", 
    x = NULL, 
    y = "Estimated effect on poverty rate (pp)"
  ) + theme_slide

ggplot(coef_df, aes(x = term, y = estimate)) + 
  geom_hline(yintercept = 0, linetype = "dashed") + 
  geom_pointrange(aes(ymin = conf.low, ymax = conf.high), size = 0.8) + 
  coord_flip() + labs(
    title = "Fixed Effects Model Coefficients", 
    subtitle = "State & Year FE; clustered by state; pop-weighted", 
    x = NULL, 
    y = "Estimated effect on poverty rate (pp)"
  ) + theme_slide
```

The fixed-effects model estimates show that, after controlling for real per capita income, real GDP, and unemployment, higher minimum wages are associated with slightly lower poverty rates across U.S.states between 2018 and 2023. The coefficient on log (min wage) is negative but small, suggesting that a 1% increase in minimum wage corresponds to an approximate 0.012 percentage point decline in the poverty rate, holding other factors constant, confirming a measurable poverty-reducing effect (albeit small). By comparison, log (income per capita) also exhibits a statistically significant and negative relationship with poverty, indicating that rising household incomes plays even a larger role in poverty alleviation. Meanwhile, both log (real GDP) and unemployment rate show small but statistically significant positive effects, suggesting that aggregate economic growth alone does not guarantee reductions in poverty, and that higher unemployment exacerbates financial hardship. Taken together, the results imply that state-level poverty reduction between 2018 and 2023 was most strongly influenced by household income gains and targeted wage policies, while broader macroeconomic growth provided less direct relief. 