---
title: "DA 401 Early Results"
author: "Calvin Burzynski"
date: "2025-10-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
library(tidyr)
library(stringr)
library(lubridate)
library(readr)
library(ggplot2)

#load in ipums (acs) dataset
ipums <- read.csv("../../../usa_00003.csv.gz")

#load minimum wage DOL data
min_wage_data <- read.csv("..//..//..//Minimum Wage from DOL 2018-2024.csv")

#load monthly unemployment rate data
unemployment_rate_data <- read.csv("..//..//..//State unemployment rates over the last 10 years, seasonally adjusted_BLS.csv")

gdp_income <- read.csv("..//..//..//State Annual Summary Statistics GDP Income.csv", skip = 3)
```

```{r}
#filter for poverty status
ipums_poverty <- ipums %>% 
  filter(EMPSTAT == 1, UHRSWORK >= 35) %>% 
  mutate(in_poverty = ifelse(INCWAGE <= 18000, 1, 0)) %>% # $18,000 threshold to account for recent changes in inflation
  group_by(STATEFIP, YEAR) %>% #group by state and year
  summarize(poverty_rate = weighted.mean(in_poverty, w = PERWT, na.rm = TRUE), .groups = "drop")
```


```{r}
#cleans columns with 'X' prefix
min_wage_cleaned <- min_wage_data %>%
  rename_with(~str_remove(., "^X"), starts_with("X"))

min_wage_long <- min_wage_cleaned %>%
  pivot_longer(
    cols = '2018':'2023', #attempted to select columns but still included 2024 column
    names_to = "YEAR",
    values_to = "min_wage"
  ) %>%
  mutate(YEAR = as.numeric(YEAR))

#drop 2024 column as it wasn't omitted when trying to pivot
min_wage_long <- min_wage_long %>%
  select(-("2024"))
```

```{r}
#pivot unemplyoment rate data to long format
str(unemployment_rate_data)

#replaces "." with "-" for month and year columns
unemployment_rate_rename <- unemployment_rate_data %>%
  rename_with(~ str_replace_all(.x, "\\.", "-"), .cols = -State)

#helps match structure of month-year columns 
month_cols <- grep("^[A-Za-z]{3}-\\d{2,4}$", names(unemployment_rate_rename), value = TRUE)

unemployment_numeric <- unemployment_rate_rename %>%
  mutate(across(all_of(month_cols), ~ parse_number(as.character(.))))

unemployment_long <- unemployment_numeric %>%
  pivot_longer(
    cols = all_of(month_cols), 
    names_to = "date_string", 
    values_to = "unemployment_rate"
  ) %>%
  mutate(
    date = my(date_string), 
    YEAR = year(date)
  ) %>%
  filter(!is.na(date), between(YEAR, 2018, 2023)) %>%
  group_by(State, YEAR) %>%
  summarise(unemployment_rate = mean(unemployment_rate, na.rm = TRUE), .groups = 'drop')
```

```{r}
gdp_income_long <- gdp_income %>%
  rename_with(~str_remove(., "^X"), starts_with("X")) %>%
  filter(LineCode %in% c(1, 4)) %>%
  select(GeoName, GeoFips, Description, '2018':'2023') %>%
  pivot_longer(
    cols = '2018':'2023',
    names_to = "YEAR",
    values_to = "value"
  ) %>%
  mutate(Year = as.numeric(YEAR)) %>%
  pivot_wider(names_from = Description, values_from = value)
```


```{r}
#lookup table to help identify by both column names
state_lookup <- min_wage_long %>%
  select(State, STATEFIP = StateFIPS) %>%
  mutate(STATEFIP = as.numeric(STATEFIP)) %>%
  distinct()

#join fips code to the unemployment data
unemployment_with_fips <- unemployment_long %>%
  left_join(state_lookup, by = "State")

gdp_income_with_fips <- gdp_income_long %>%
  left_join(state_lookup, by = c("GeoName" = "State"))



#needed to make sure data types align
min_wage_long <- min_wage_long %>%
  mutate(YEAR = as.numeric(YEAR))

gdp_income_with_fips <- gdp_income_with_fips %>% mutate(YEAR = as.numeric(YEAR))

unemployment_with_fips <- unemployment_with_fips %>% mutate(YEAR = as.numeric(YEAR))

ipums_poverty <- ipums_poverty %>% mutate(YEAR = as.numeric(YEAR))

#initial panel data
initial_panel <- ipums_poverty %>%
  left_join(min_wage_long, by = c("STATEFIP" = "StateFIPS", "YEAR")) %>% #add minimum wage data
  left_join(unemployment_with_fips, by = c("STATEFIP", "YEAR")) %>% #add unemployment data
  left_join(gdp_income_with_fips, by = c("STATEFIP", "YEAR"))
```

```{r}
plot_data <- initial_panel %>%
  group_by(State.x) %>%
  mutate(Group = ifelse(any(min_wage > 7.25), "Treated States", "Control States")) %>%
  ungroup()

ggplot(plot_data, aes(x = YEAR, y = poverty_rate, color = Group)) + 
  stat_summary(fun = "mean", geom = "line", size = 1.2) + 
  stat_summary(fun = "mean", geom = "point", size = 3) + 
  labs(title = "Poverty Rate Trends for Low-Wage Workers (2018-2023)", 
       subtitle = "Visual check for the parallel trends assumption", 
       x = "Year", 
       y = "Average Poverty Rate", 
       color = "State Group") + 
  theme_minimal()
```

The yielded plot from the code is a visualization check for the parallel trends assumption. For Difference-in-Differences to be valid, it must be assumed that the "Treated" and "Control" states would have followed similar trends if no minimum wage increases had occurred. The plot essentially shows this. A large spike in poverty for both groups in 2020 is a clear and expected result, potentially reflecting the economic disruption of the covid-19 pandemic. After 2020/21, there appeared to be a decline in poverty for both groups. The hope is that the analysis going forward will answer the following question: "did the poverty rate in the treated states fall more than it did in the control states, after accounting for the baseline differences and 2020 shock?"